Using structural knowledge to correct RNA pyrosequencing errors in non-coding RNAs

### 2013-05-20 JCB/methods_JCB.tex
New inside/outside figures
Started adapting text

Let us then remind the Hamming distance function $\delta: B^*\times B^* \to \mathbb{N}^+$, which takes two sequences $s'$ and $s''$ as input, $|s'|=|s''|$, and returns the number of differing positions.
Finally, let us denote by $E_{(i,j),ab \to ab'}^{\Omega,\beta}$ the local contribution of a base-pair $(i,j)$ to the pseudo-energy, such that
\begin{equation}
  E_{(i,j),ab \to a'b'}^{\Omega,\beta}  = \alpha \cdot\ES^\beta_{a b \to a' b'}+(1-\alpha)\cdot\EI^{\Omega}_{(i,j),a'b'}.
\end{equation}
\subsubsection{Inside computation.}
\begin{figure}[t]
\resizebox{\textwidth}{!}{\input{FigDPInside.pgf}}
\caption{Principle of the inside computation (partition function). Any sequence with mutations  
can be decomposed as a sequence preceded by a, possibly mutated, base 
(Unpaired case), a sequence surrounded by some base-pair (Stacking-pair case), 
or as two sequences segregated by some base-pair (General base-pairing case), and mutations must be distributed between sub-sequences 
and instantiated bases.}
\end{figure}

The \emph{Inside} function $\Z{i,j}{m}{a,b}$ is the partition function, i.e. the sum of Boltzmann factors over all sequences in the interval $[i,j]$, at distance $m$ of $s_{[i,j]}$, and having flanking nucleotides $a$ and $b$ (at positions $i-1$ and $j+1$ respectively). 
\TODOYann{Reformat inside Figure as in Incarnation.}
Such terms can be defined by recurrence, for which the following initial conditions holds:

\begin{equation}
	\forall i \in [0,n-1]:\, \Z{i+1,i}{m}{a,b}=\left\{
	\begin{array}{ll}
		1 &\text{If } m = 0\\
		0 &\text{Otherwise.}
	\end{array}\right.
\label{eq:Z_in}
\end{equation}
In other words, the set of sequences at distance $m$ of the empty sequence is either empty if $m>0$, or restricted to the empty sequence, having energy $0$, if $m=0$. Since the energetic terms only depend on base pairs, they are not involved in the initial conditions. 
The main recursion itself is composed of four terms:\\

\begin{equation}
\label{eq:Z_rec}
\resizebox{.93\textwidth}{!}{%
	$\Z{i,j}{m}{a,b}:=\left\{
  \begin{array}{l}
  		\displaystyle
      \sum_{\substack{a'\in \B,\\ \Kron_{a',s_i}\le m}}  
      \Z{i+1,j}{m-\Kron_{a',s_i}}{a',b} \\
       \text{[If }S_{i}=-1\text{]}\\
      \displaystyle
      \sum_{\substack{a',b'\in \B^2,\\ \Kron_{a'b',s_is_j}\le m}}
			 e^{\frac{-E_{(i,j),ab \to a'b'}^{\Omega,\beta}}{RT}}
			 \cdot \Z{i+1,j-1}{m-\Kron_{a'b',s_is_j}}{a',b'}\\
			 \text{[Else if }S_i=j \land S_{i-1}=j+1\text{]}\\
			 \displaystyle
      \sum_{\substack{a',b'\in \B^2,\\ \Kron_{a'b',s_is_k}\le m}}
      \sum_{m'=0}^{m-\Kron_{a'b',s_is_k}}
   		 e^{\frac{-E_{(i,k),\varnothing\to a'b'}^{\Omega,\beta}}{RT}}
      \cdot\Z{i+1,k-1}{m-\Kron_{a'b',s_is_k}-m'}{a',b'}
      \cdot\Z{k+1,j}{m'}{b',b} \\
       \text{[Elif }S_i=k \land i < k \leq j\text{]}\\
      0 \\\text{[Otherwise]}
	\end{array}\right.$}
\end{equation}
\TODOYann{Rephrase in term of tree, as in Incarnation??}
The cases can be broken down as follows:
\begin{description}
\item[$S_{i}=-1$:] If nucleotide at position $i$ is unpaired, then 
any sequence is a concatenation of a, possibly mutated, nucleotide $a'$ at position  $i$, 
followed by a sequence over $[i+1,j]$ having $m-\Kron_{a',s_i}$ mutations (accounting for a possible mutation 
at position $i$), and having flanking nucleotides $a'$ and $b$.
\item[$S_i=j$ and $S_{i-1}=j+1$:] Any sequence generated in $[i,j]$ consists of two, possibly mutated, nucleotides $a'$ and $b'$, flanking a sequence over $[i+1,j-1]$ having distance $m-\Kron_{a'b',s_is_j}$.
Since positions $i$ and $i-1$ are paired with $j$ and $j+1$ respectively, 
then a stacking energy contribution is added. 
\item[$S_i=k$ and $i<k \leq j$:] If position $i$ is paired and not involved in a stacking, then the 
only term contributing directly to the energy is the isostericity of the base pair $(i,k)$. 
Any sequence on $[i,j]$ consists of two nucleotide $a'$ and $b'$ at positions $i$ and $k$ respectively, flanking a sequence over interval $[i+1,k-1]$ and preceding a (possibly empty) sequence interval $[k+1,j]$. Since the number of mutations sum to $m$ over the whole sequence must , then a parameter $m'$ is introduced to distribute the remaining mutations between the two sequences.
\item[Else:] In any other case, we are in a derivation of the SCFG that does not correspond to the secondary structure $S$, and we return $0$.
\end{description}

\subsubsection{Outside computation.}	
\begin{figure}[t]
\begin{minipage}{0.79\textwidth}
\resizebox{\textwidth}{!}{\input{FigDPOutside.pgf}}
\end{minipage}
\begin{minipage}{0.20\textwidth}
\caption{Principle of the outside computation. Note that the outside algorithm uses intermediate results from the inside algorithm, 
therefore its efficient implementation requires a precomputation of the inside contributions.}
\end{minipage}
\SpaceCheating
\end{figure}

>>>>

Let us then remind the Hamming distance function $\delta: B^*\times B^* \to \mathbb{N}^+$, which takes two sequences $s'$ and $s''$ as input, $|s'|=|s''|$, and returns the number of differing positions.
Finally, let us denote by $E_{(i,j),ab \to ab'}^{\Omega,\beta}$ the local contribution of a base-pair $(i,j)$ to the pseudo-energy, such that
\begin{equation}
  E_{(i,j),ab \to a'b'}^{\Omega,\beta}  = \alpha \cdot\ES^\beta_{a b \to a' b'}+(1-\alpha)\cdot\EI^{\Omega}_{(i,j),a'b'}.
\end{equation}
\subsubsection{Inside computation.}
\begin{figure}[t]\centering
\includegraphics[scale=\ScaleDP]{FigDPInsideWrapper}
\caption{Principle of the inside computation (partition function). Any sequence with mutations  
can be decomposed as a sequence preceded by a, possibly mutated, base 
(Unpaired case), a sequence surrounded by some base-pair (Stacking-pair case), 
or as two sequences segregated by some base-pair (General base-pairing case), and mutations must be distributed between sub-sequences 
and instantiated bases.}
\end{figure}

The \emph{Inside} function $\Z{\Struct}{m}{x,y}$ is simply the partition function, i.e. the sum of Boltzmann factors over all sequences for a substructure $\Struct$, featuring $m$ mutations/errors compared to $s$, and having flanking nucleotides $x$ and $y$. 
\TODOYann{Reformat inside Figure as in Incarnation.}
Such terms can be computed by recurrence, whose initial values are:
\begin{equation}
	\forall x,y\in B\times B, m\in [0,M]:\, \Z{\varepsilon}{m}{x,y}=\left\{
	\begin{array}{ll}
		1 &\text{If } m = 0\\
		0 &\text{Otherwise.}
	\end{array}\right.
\label{eq:Z_in}
\end{equation}
In other words, either there is no sequence at distance $m>0$ of the empty sequence, or the only allowed sequence is the empty sequence ($m=0$), having energy $0$. Since the energetic terms only depend on base pairs, they are not involved in the initial conditions. 
The main recursion itself is composed of four terms:\\

\begin{equation}
\label{eq:Z_rec}
	\Z{i,j}{m}{a,b}:=\left\{
  \begin{array}{ll}
  		\displaystyle
      \sum_{\substack{a'\in \B,\\ \Kron_{a',s_i}\le m}}  
      \Z{i+1,j}{m-\Kron_{a',s_i}}{a',b} &
       \text{[If }S_{i}=-1\text{]}\\
      \displaystyle
      \sum_{\substack{a',b'\in \B^2,\\ \Kron_{a'b',s_is_j}\le m}}
			 e^{\frac{-E_{(i,j),ab \to a'b'}^{\Omega,\beta}}{RT}}
			 \cdot \Z{i+1,j-1}{m-\Kron_{a'b',s_is_j}}{a',b'}&
			 \text{[Else if }S_i=j \land S_{i-1}=j+1\text{]}\\
			 \displaystyle
      \sum_{\substack{a',b'\in \B^2,\\ \Kron_{a'b',s_is_k}\le m}}
      \sum_{m'=0}^{m-\Kron_{a'b',s_is_k}}
   		 e^{\frac{-E_{(i,k),\varnothing\to a'b'}^{\Omega,\beta}}{RT}}
      \cdot\Z{i+1,k-1}{m-\Kron_{a'b',s_is_k}-m'}{a',b'}
      \cdot\Z{k+1,j}{m'}{b',b} &
       \text{[Elif }S_i=k \land i < k \leq j\text{]}\\
      0 \\\text{[Otherwise]}
	\end{array}\right.
\end{equation}
\TODOYann{Rephrase in term of tree, as in Incarnation??}
The cases can be broken down as follows:
\begin{description}
\item[$S_{i}=-1$:] If nucleotide at position $i$ is unpaired, then 
any sequence is a concatenation of a, possibly mutated, nucleotide $a'$ at position  $i$, 
followed by a sequence over $[i+1,j]$ having $m-\Kron_{a',s_i}$ mutations (accounting for a possible mutation 
at position $i$), and having flanking nucleotides $a'$ and $b$.
\item[$S_i=j$ and $S_{i-1}=j+1$:] Any sequence generated in $[i,j]$ consists of two, possibly mutated, nucleotides $a'$ and $b'$, flanking a sequence over $[i+1,j-1]$ having distance $m-\Kron_{a'b',s_is_j}$.
Since positions $i$ and $i-1$ are paired with $j$ and $j+1$ respectively, 
then a stacking energy contribution is added. 
\item[$S_i=k$ and $i<k \leq j$:] If position $i$ is paired and not involved in a stacking, then the 
only term contributing directly to the energy is the isostericity of the base pair $(i,k)$. 
Any sequence on $[i,j]$ consists of two nucleotide $a'$ and $b'$ at positions $i$ and $k$ respectively, flanking a sequence over interval $[i+1,k-1]$ and preceding a (possibly empty) sequence interval $[k+1,j]$. Since the number of mutations sum to $m$ over the whole sequence must , then a parameter $m'$ is introduced to distribute the remaining mutations between the two sequences.
\item[Else:] In any other case, we are in a derivation of the SCFG that does not correspond to the secondary structure $S$, and we return $0$.
\end{description}

\subsubsection{Outside computation.}	
\begin{figure}[t]\centering
\includegraphics[scale=\ScaleDP]{FigDPOutsideWrapper}
\caption{Principle of the outside computation. Note that the outside algorithm uses intermediate results from the inside algorithm, 
therefore its efficient implementation requires a precomputation of the inside contributions.}
\end{figure}
